---
title: "code_only"
author: "Yaelle Tang"
date: "2026-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
#Load libraries
#install.packages("rio")
library(tidyverse)
library(readxl)
library(janitor)
library(rio)
```



```{r chart-gdp-growth echo=TRUE}
# Retrieve GDP Data
gdp_county <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/bea_gdp_2024_county_CAGDP1.csv", skip=3) |> 
  clean_names() |> 
    mutate(across(3:5, as.numeric))

# Calculate change 2024 vs 2022; split out state
gdp_county <- gdp_county |> 
  mutate(pct_24_v_22 = (x2024-x2022)/x2022,
          state = str_split_fixed(geo_name, ", ", 2)[, 2],
         county = str_split_fixed(geo_name, ", ", 2)[, 1])  

philly_gdp <- gdp_county |>
  filter(state == "PA" | state == "NJ") |>
  filter(county %in% c("Philadelphia", "Montgomery", 
                        "Bucks", "Chester", 
                        "Delaware", "Gloucester", 
                        "Camden", "Burlington")) |>
  select(county, state, x2022, x2024, pct_24_v_22) |>
  mutate(rank(x2022)) |>
  mutate(rank(x2024)) |>
  arrange(desc(pct_24_v_22))

# Chart: GDP Growth by County
  # AI assisted

ggplot(philly_gdp, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "GDP Growth by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()

```


::: {.callout-note collapse="true"}
## About the data
BEA CAINC1 - County Personal Income Summary 
CAINC1 County personal income summary: personal income, population, per capita personal income
Source: [Personal income (Thousands of dollars)](https://www.bea.gov/news/2026/gross-domestic-product-county-and-personal-income-county-2024)
:::

```{r}
#Load and Clean Data
cainc1 <- read_csv("../data/bea_gdp_2024_personal_income_county_CAINC1.csv", skip=3) |> 
  clean_names() |> 
  mutate(across(x2022:x2024, as.numeric)) |> 
  mutate(
    pct_23_v_22 = round((x2023 - x2022) / x2022, 4),
    pct_24_v_23 = round((x2024 - x2023) / x2023, 4),
    pct_24_v_22 = round((x2024 - x2022) / x2022, 4)
  )

#Filter Greater Philadelphia
philly_income <- cainc1 |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ")) |>
  select(geo_name, x2022, x2024, pct_24_v_22) |>
  separate(geo_name, into = c("county", "state"), sep = ", ") |>
  arrange(desc(pct_24_v_22))

#Chart: Personal Income Growth by County
  #AI assisted

ggplot(philly_income, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Personal Income Growth by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()
```
*BEA CAGDP2 — GDP by Industry (Current dollars)* ![CAGDP2 GDP by industry in current dollars](https://apps.bea.gov/itable/?ReqID=70&step=1&_gl=1*ov6bhy*_ga*OTk4Njc0OTc1LjE3ODAyODAzNTY.*_ga_J4698JNNFT*czE3ODAzNDIwNjgkbzQkZzEkdDE3ODAzNDIxMzIkajYwJGwwJGgw#eyJhcHBpZCI6NzAsInN0ZXBzIjpbMSwyOSwyNSwzMV0sImRhdGEiOltbIlRhYmxlSWQiLCI1MDEiXSxbIk1ham9yX0FyZWEiLCI0Il1dfQ==)

```{r}
# Load and Clean Data
# Merge Pennsylvania and New Jersey industry files
gdp_industry <- bind_rows(
  read_csv("../data/CAGDP2_GPD_PA_Industry.csv", skip = 3) |> clean_names(),
  read_csv("../data/CAGDP2_GPD_NJ_Industry.csv", skip = 3) |> clean_names()
)

# Using Private industries sub-categories only (excluding aggregates to avoid double counting)
gdp_industry <- gdp_industry |>
 mutate(
  x2022 = as.numeric(x2022),
  x2023 = as.numeric(x2023),
  x2024 = as.numeric(x2024)
)

# AI-assisted
greater_philly_total <- gdp_industry |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ")) |>
  filter(line_code == 1) |> 
  select(geo_name, description, x2022, x2023, x2024) |>
  mutate(pct_24_v_22 = round((x2024 - x2022) / x2022, 4)) |>
  arrange(desc(pct_24_v_22))

greater_philly_industry <- gdp_industry |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ"))  |>
  filter(description %in% c(
    "Agriculture, forestry, fishing and hunting",
    "Mining, quarrying, and oil and gas extraction",
    "Utilities",
    "Construction",
    "Manufacturing",
    "Wholesale trade",
    "Retail trade",
    "Transportation and warehousing",
    "Information",
    "Finance, insurance, real estate, rental, and leasing",
    "Professional and business services",
    "Educational services, health care, and social assistance",
    "Arts, entertainment, recreation, accommodation, and food services",
    "Other services (except government and government enterprises)"
  )) |>
  select(geo_name, description, x2022, x2024) |>
  mutate(pct_24_v_22 = round((x2024 - x2022) / x2022, 4)) |>
  mutate(x24_v_22 = x2024 - x2022) |>
  arrange(desc(x24_v_22))

greater_philly_industry |>
  slice_min(x24_v_22, n = 10) |>
  select(geo_name, description, x24_v_22)

delaware_gdp_industry <- greater_philly_industry |>
  filter(geo_name == "Delaware, PA") |>
  filter(x2022 > 0) |>
  mutate(abs_change = x2024 - x2022) |>
  arrange(abs_change)

delaware_gdp_industry |>
  select(geo_name, description, x24_v_22)


```

> **Finding:** Delaware County's manufacturing sector ranks third on the list, which likely explains why it is the only county in the region with negative GDP growth. I decide to have a closer look at Delaware County's industry breakdown.

```{r}
philly_manufacturing <- greater_philly_industry |>
  filter(description == "Manufacturing") |>
  filter(x2022 > 0) |> #AI-assisted
  mutate(abs_change = x2024 - x2022) |>
  separate(geo_name, into = c("county", "state"), sep = ", ") |>
  arrange(abs_change)

#AI-assister
ggplot(philly_manufacturing, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Manufacturing GDP Percentage Change by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()

ggplot(philly_manufacturing, aes(x = reorder(county, abs_change), y =abs_change , fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Manufacturing GDP Change by County in Greater Philadelphia (2022-2024)",
    subtitle = "In thousands of dollars",
    x = NULL,
    y = "Absolute Change",
    fill = "State"
  ) +
  theme_minimal()
```
> **Findings:** Philadelphia had the largest manufacturing decline both in percentage terms and absolute value, but its overall GDP remained positive. Delaware County, by contrast, lost 12% of its manufacturing output and saw its overall GDP turn negative. Maybe its economy is more vulnerable and lacks other industries to fill the gap left by manufacturing's decline.