---
title: "Economic portrait of greater Philadelphia"
author: 
  - name: Yaelle Tang
    email: yueyang_tang@berkeley.edu
date: "2026-06-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

**Rsw comment:** Yaelle - You've done a tremendous amount of work and
carefully followed my guidance from yesterday. This project is in good
shape at this point. It is also quite long so we want to find ways to
cut it back and emphasize the main findings. I put a few comments under
**rsw comment** below.

# Pitch Memo

## In two sentences, what is this story about?

**rsw comment** - great focus on an important trend. really captures
your attention

Delaware County is the only county in the greater Philadelphia region
whose GDP shrank between 2022 and 2024. Manufacturing industry in
**spelling \\ Dalaware** County lost more than \$428 million in output,
showing a 12% decline. Meanwhile, neighboring counties like Bucks and
Burlington saw their manufacturing GDP grow by double digits over the
same period.

## In one sentence, why tell this story NOW?

The Bureau of Economic Analysis (BEA) released its 2024 county-level GDP
data on February 5, 2026, offering the most current picture of how the
greater Philadelphia region's economy has shifted.

## Who is your target audience?

Residents and workers in the greater Philadelphia area

## Three people you can interview for this story

Working on it...

## Potential impact

A sustained decline in Delaware County's manufacturing output likely
means fewer stable jobs for working-class residents, reduced tax revenue
for local services, and a widening gap between the county and its
neighbors.

## What are three things that surprised or interested you about this story?

**rsw comment** great observation. let's show this in a chart below.

1.  Delaware County's GDP shrank even as personal income grew by 10%.
    Residents may be earning more by commuting to jobs elsewhere while
    local economic activity continues to decline. How to explore this
    with data?
2.  Philadelphia's manufacturing sector actually fell harder than
    Delaware County's (14% vs. 12%), but Philadelphia's overall GDP
    remained positive. Delaware County's economy is far more dependent
    on manufacturing and has fewer industries to fill the gap.
3.  Manufacturing's share of the economy is shrinking across nearly
    every county in greater Philadelphia.

**rsw comment: how does the decline in manufacturing compare to national
trends? it will be on the BEA GDP press release**

## Context: Summarize any previous coverage.

[WHYY reported in
2023](https://whyy.org/articles/manufacturing-downturn-economy-federal-reserve/)
that the greater Philadelphia region experienced nine consecutive months
of manufacturing slowdown, with factory closures and layoffs in
Philadelphia and King of Prussia. But previous coverage treated the
region as a whole. No reporting has specifically asked why Delaware
County, alone among its neighbors, failed to compensate for
manufacturing losses with growth in other sectors

## Photos, video, graphics - How will you make this story visual?

**rsw comments: these charts aren't displaying properly**

![Manufacturing GDP Change by County in Greater
Philadelphia](../img/GDP_Growth_by_County.png)

![Manufacturing GDP Change by County in Greater
Philadelphia](../img/Manufacturing_GDP_Change.png) ![Percentage
change](../img/Manufacturing_GDP_Percentage_Change.png)

## What other information do you need to gather?

-   Employment data for Delaware County to understand where workers went
    as manufacturing shrank, and why personal income continued to rise
    despite the GDP decline.
-   Population change data would help confirm whether residents are
    leaving the county.
-   Any specific factory closures during this period?
-   A deeper breakdown of manufacturing sub-sectors would identify which
    specific industries drove the decline.

## Estimated delivery

# Code Memo

**Data limitations**

This data series lags significantly. We are working with 2024 data, the
most recent available. But it is still the best you can get and you
can't beat the price.

## Setup

```{r include=FALSE}
#Load libraries
#install.packages("rio")
library(tidyverse)
library(readxl)
library(janitor)
library(rio)
```

## Regional GDP by County

**Data limitations**

This data series lags significantly. We are working with 2024 data, the
most recent available. But it is still the best you can get and you
can't beat the price.

<br>

### Retrieve GDP Data

GDP CAGDP1 County gross domestic product (GDP) summary Real Gross
Domestic Product (GDP) (Thousands of chained 2017 dollars) Source Page
https://www.bea.gov/news/2026/gross-domestic-product-county-and-personal-income-county-2024

And more detail: https://www.bea.gov/data/gdp/gdp-by-county

```{r}
gdp_county <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/bea_gdp_2024_county_CAGDP1.csv", skip=3) |> 
  clean_names() |> 
    mutate(across(3:5, as.numeric))

glimpse(gdp_county)
```

### Calculate change 2024 vs 2022; split out state

```{r}

gdp_county <- gdp_county |> 
  mutate(pct_24_v_22 = (x2024-x2022)/x2022,
          state = str_split_fixed(geo_name, ", ", 2)[, 2],
         county = str_split_fixed(geo_name, ", ", 2)[, 1])  

```

```{r}
philly_gdp <- gdp_county |>
  filter(state == "PA" | state == "NJ") |>
  filter(county %in% c("Philadelphia", "Montgomery", 
                        "Bucks", "Chester", 
                        "Delaware", "Gloucester", 
                        "Camden", "Burlington")) |>
  select(county, state, x2022, x2024, pct_24_v_22) |>
  mutate(rank(x2022)) |>
  mutate(rank(x2024)) |>
  arrange(desc(pct_24_v_22))

philly_gdp
```

### Chart: GDP Growth by County
**rsw comment: you'll want this chart at the top of your page**
```{r chart-gdp-growth}
# AI assisted
ggplot(philly_gdp, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "GDP Growth by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()
```

> **Finding:** Delaware County was the only county in the region to see
> its GDP shrink between 2022 and 2024.

## Part 2: Personal Income by County

### Load and Clean Data

```{r}
# [Personal income (Thousands of dollars)](https://www.bea.gov/news/2026/gross-domestic-product-county-and-personal-income-county-2024)
# BEA CAINC1 - County Personal Income Summary
# CAINC1 County personal income summary: personal income, population, per capita personal income
cainc1 <- read_csv("../bea_gdp_2024_personal_income_county_CAINC1.csv", skip=3) |> 
  clean_names() |> 
  mutate(across(x2022:x2024, as.numeric)) |> 
  mutate(
    pct_23_v_22 = round((x2023 - x2022) / x2022, 4),
    pct_24_v_23 = round((x2024 - x2023) / x2023, 4),
    pct_24_v_22 = round((x2024 - x2022) / x2022, 4)
  )
```

### Filter Greater Philadelphia

```{r}
philly_income <- cainc1 |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ")) |>
  select(geo_name, x2022, x2024, pct_24_v_22) |>
  separate(geo_name, into = c("county", "state"), sep = ", ") |>
  arrange(desc(pct_24_v_22))

philly_income
```

### Chart: Personal Income Growth by County

```{r}
# AI assisted
ggplot(philly_income, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Personal Income Growth by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()

```

> **Finding:** Personal income in Delaware County grew 10.98% between
> 2022 and 2024, yet its GDP was the only one in the region to shrink.
> Residents may be earning more by working elsewhere while local
> industries are losing ground.

## Part 3: GDP by Industry

### Load and Clean Data

*BEA CAGDP2 — GDP by Industry (Current dollars)* ![CAGDP2 GDP by
industry in current
dollars](https://apps.bea.gov/itable/?ReqID=70&step=1&_gl=1*ov6bhy*_ga*OTk4Njc0OTc1LjE3ODAyODAzNTY.*_ga_J4698JNNFT*czE3ODAzNDIwNjgkbzQkZzEkdDE3ODAzNDIxMzIkajYwJGwwJGgw#eyJhcHBpZCI6NzAsInN0ZXBzIjpbMSwyOSwyNSwzMV0sImRhdGEiOltbIlRhYmxlSWQiLCI1MDEiXSxbIk1ham9yX0FyZWEiLCI0Il1dfQ==)

```{r}
# Merge Pennsylvania and New Jersey industry files
gdp_industry <- bind_rows(
  read_csv("../CAGDP2_GPD_PA_Industry.csv", skip = 3) |> clean_names(),
  read_csv("../CAGDP2_GPD_NJ_Industry.csv", skip = 3) |> clean_names()
)

glimpse(gdp_industry)
```

### Filter Private Industry Sub-sectors for Greater Philadelphia

```{r}
# Using Private industries sub-categories only (excluding aggregates to avoid double counting)
gdp_industry <- gdp_industry |>
 mutate(
  x2022 = as.numeric(x2022),
  x2023 = as.numeric(x2023),
  x2024 = as.numeric(x2024)
)

# AI-assisted
greater_philly_total <- gdp_industry |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ")) |>
  filter(line_code == 1) |> 
  select(geo_name, description, x2022, x2023, x2024) |>
  mutate(pct_24_v_22 = round((x2024 - x2022) / x2022, 4)) |>
  arrange(desc(pct_24_v_22))

glimpse(greater_philly_total)

# check what industry included
gdp_industry|>
  distinct(description) |>
  print(n = 36)
```

```{r}
greater_philly_industry <- gdp_industry |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ"))  |>
  filter(description %in% c(
    "Agriculture, forestry, fishing and hunting",
    "Mining, quarrying, and oil and gas extraction",
    "Utilities",
    "Construction",
    "Manufacturing",
    "Wholesale trade",
    "Retail trade",
    "Transportation and warehousing",
    "Information",
    "Finance, insurance, real estate, rental, and leasing",
    "Professional and business services",
    "Educational services, health care, and social assistance",
    "Arts, entertainment, recreation, accommodation, and food services",
    "Other services (except government and government enterprises)"
  )) |>
  select(geo_name, description, x2022, x2024) |>
  mutate(pct_24_v_22 = round((x2024 - x2022) / x2022, 4)) |>
  mutate(x24_v_22 = x2024 - x2022) |>
  arrange(desc(x24_v_22))

glimpse(greater_philly_industry)
```

### Top 10 Industries with Largest Decline Across the Region

```{r}
greater_philly_industry |>
  slice_min(x24_v_22, n = 10)
```

> **Finding:** Delaware County's manufacturing sector ranks third on the
> list, which likely explains why it is the only county in the region
> with negative GDP growth. I decide to have a closer look at Delaware
> County's industry breakdown.

### Which Industries Are Shrinking in Delaware County

Focus on Delaware County to identify the main driver of GDP decline

```{r}
delaware_gdp_industry <- greater_philly_industry |>
  filter(geo_name == "Delaware, PA") |>
  filter(x2022 > 0) |>
  mutate(abs_change = x2024 - x2022) |>
  arrange(abs_change)

delaware_gdp_industry
```

> **Finding:** Manufacturing is the primary driver of Delaware County's
> GDP decline, with output falling 12.3% and losing over \$428 million
> between 2022 and 2024.

### Manufacturing Across Greater Philadelphia

```{r}
philly_manufacturing <- greater_philly_industry |>
  filter(description == "Manufacturing") |>
  filter(x2022 > 0) |> #AI-assisted
  mutate(abs_change = x2024 - x2022) |>
  separate(geo_name, into = c("county", "state"), sep = ", ") |>
  arrange(abs_change)

philly_manufacturing
```

### Manufacturing Share of Total GDP by County

Compare manufacturing's share of each county's economy in 2022 vs 2024

```{r}
gdp_industry |>
  filter(str_detect(geo_name, "Philadelphia, PA|Montgomery, PA|Bucks, PA|Chester, PA|Delaware, PA|Gloucester, NJ|Camden, NJ|Burlington, NJ")) |>
  filter(description %in% c("All industry total", "Manufacturing")) |>
  filter(x2022 > 0) |>
  select(geo_name, description, x2022, x2024) |>
  pivot_wider(names_from = description, values_from = c(x2022,x2024)) |>
  mutate(
    manufacturing_share_22 = round(x2022_Manufacturing / `x2022_All industry total`, 4),
    manufacturing_share_24 = round(x2024_Manufacturing / `x2024_All industry total`, 4),) |>
  separate(geo_name, into = c("county", "state"), sep = ", ") |>
  arrange(desc(manufacturing_share_22))
```

> **Findings:** Manufacturing's share of GDP declined in nearly every
> county across greater Philadelphia between 2022 and 2024.
> Manufacturing in Delaware County is shrinking fastest.

### Charts: Manufacturing GDP Change by County

```{r}
#AI-assister
ggplot(philly_manufacturing, aes(x = reorder(county, pct_24_v_22), y = pct_24_v_22, fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Manufacturing GDP Percentage Change by County in Greater Philadelphia (2022-2024)",
    x = NULL,
    y = "Percent Change",
    fill = "State"
  ) +
  theme_minimal()

ggplot(philly_manufacturing, aes(x = reorder(county, abs_change), y =abs_change , fill = state)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Manufacturing GDP Change by County in Greater Philadelphia (2022-2024)",
    subtitle = "In thousands of dollars",
    x = NULL,
    y = "Absolute Change",
    fill = "State"
  ) +
  theme_minimal()
```

> **Findings:** Philadelphia had the largest manufacturing decline both
> in percentage terms and absolute value, but its overall GDP remained
> positive. Delaware County, by contrast, lost 12% of its manufacturing
> output and saw its overall GDP turn negative. Maybe its economy is
> more vulnerable and lacks other industries to fill the gap left by
> manufacturing's decline.
